---
title: "NOAA Weather Damage Analysis"
output:
  html_document:
    fig_caption: yes
    keep_md: yes
    toc: yes
  html_notebook: default
  pdf_document:
    toc: yes
---


```{r setup, cache = TRUE, echo=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	include = FALSE
)
library(knitr)
library(tidyverse)
library(lubridate)
library(stringr)
library(RCurl)
require(reshape2)

```
# Reading Data 

Event information was taken from the  [NOAA](ftp://ftp.ncdc.noaa.gov/pub/data/swdi/stormevents/csvfiles/) database of storm events. The events are downloaded from NOAA when this report is run so that we can get the most recently updated data. 

```{r main_data_frame, message=FALSE, warning=FALSE, cache=TRUE, include=FALSE}

#create download and save-file URLs
# To analyze most recent data use
url <-        "ftp://ftp.ncdc.noaa.gov/pub/data/swdi/stormevents/csvfiles/" 
# To speed up for testing, ftp docs to "data/" and uncomment the next line

# url <- "data/"

dld_base <- "StormEvents_details-ftp_v1.0_d"

if (!file.exists("data")) {
        dir.create("data")
}

if (!exists("wedf")) { # If not, get to work
# To use downloaded files, use the next line to parse only
        #filenames <- dir("data/") 
# To parse files from NWS, use the next line and comment out above
         filenames <-  getURL(url, ftp.use.epsv = FALSE, dirlistonly = TRUE) 
        filenames <- strsplit(filenames, "\r\n")
        filenames = unlist(filenames)

# detail.files contains the list of the files that we actually 
# want to download 

detail.files <- filenames[grep(dld_base, filenames)]

        filenames = ""
        for (i in 1997:2016) {
                filenames[i - 1996] <- detail.files[grep(paste0(".*_d",i,"_.*"),detail.files)]
        }

        wedf <- data.frame()
        # read files into Weather Event Data Frame (wedf)
        for(filename in filenames) {
         message("downloading and parsing ", filename)        
         wedf <- bind_rows(wedf,
                read_csv(file = (paste0(url, filename))
                          , col_types = 
                                cols(BEGIN_YEARMONTH = "i",
                                 BEGIN_DAY = "i",
                                 BEGIN_TIME = "c",
                                 END_YEARMONTH = "i",
                                 END_DAY = "i",
                                 END_TIME = "c",
                                 EPISODE_ID = "i",
                                 EVENT_ID = "i",
                                 STATE = "c",
                                 STATE_FIPS = "i",
                                 YEAR = "i",
                                 MONTH_NAME = "c",
                                 EVENT_TYPE = "c",
                                 CZ_TYPE = "c",
                                 CZ_FIPS = "i",
                                 CZ_NAME = "c",
                                 WFO = "c",
                                 BEGIN_DATE_TIME = "c",
                                 CZ_TIMEZONE = "c",
                                 END_DATE_TIME = "c",
                                 INJURIES_DIRECT = "i",
                                 INJURIES_INDIRECT = "i",
                                 DEATHS_DIRECT = "i",
                                 DEATHS_INDIRECT = "i",
                                 DAMAGE_PROPERTY = "c",
                                 DAMAGE_CROPS = "c",
                                 SOURCE = "c",
                                 MAGNITUDE = "d",
                                 MAGNITUDE_TYPE = "c",
                                 FLOOD_CAUSE = "c",
                                 CATEGORY = "d",
                                 TOR_F_SCALE = "c",
                                 TOR_LENGTH = "d",
                                 TOR_WIDTH = "d",
                                 TOR_OTHER_WFO = "c",
                                 TOR_OTHER_CZ_STATE = "c",
                                 TOR_OTHER_CZ_FIPS = "i",
                                 TOR_OTHER_CZ_NAME = "c",
                                 BEGIN_RANGE = "i",
                                 BEGIN_AZIMUTH = "c",
                                 BEGIN_LOCATION = "c",
                                 END_RANGE = "i",
                                 END_AZIMUTH = "c",
                                 END_LOCATION = "c",
                                 BEGIN_LAT = "d",
                                 BEGIN_LON = "d",
                                 END_LAT = "d",
                                 END_LON = "d",
                                 EPISODE_NARRATIVE = "c",
                                 EVENT_NARRATIVE = "c",
                                 DATA_SOURCE = "c"
                          )              
                )
         )
        }

}

head(wedf)
```
Then, we will find the types of weather events causing the highest number of injuries and fatalities over the past twenty years. 

```{r subsetting, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
wetbl <- as.tbl(
        select(wedf
        , id = EVENT_ID
        , st = STATE_FIPS
        , cz = CZ_FIPS
        , type = EVENT_TYPE
        , begin = BEGIN_DATE_TIME
        , tz = CZ_TIMEZONE
        , INJURIES_DIRECT:DAMAGE_CROPS
        , fscale = TOR_F_SCALE
        , begin_lat = BEGIN_LAT
        , begin_lon = BEGIN_LON 
        , end_lat = END_LAT
        , end_lon = END_LON
        )) %>%
        mutate(type = tolower(type)) %>%
        mutate(type = gsub("heavy wind", "high wind", type)) %>%
        mutate(type = gsub("high snow", "heavy snow", type)) %>%
        mutate(type = gsub("^hurricane$", "hurricane (typhoon)", type)) %>%
        mutate(type = gsub("landslide", "avalanche", type)) %>%
        mutate(type = gsub("thunderstorm winds?.*", "thunderstorm wind", type)) %>%
        mutate(type = gsub("volcanic ashfall", "volcanic ash", type)) %>%
        mutate(type = gsub("tornado/waterspout", "waterspout", type)) %>%
        mutate(type = str_to_title(type)) %>%
        mutate(fscale = as.factor(gsub("^E?F","F",fscale))) %>%
        dplyr::filter(type != "Northern Lights") %>%
        dplyr::filter(type != "Other") %>%
        mutate( type = as.factor(type)
                ,st = as.factor(sprintf("%02d",st))
                ,cz = as.factor(sprintf("%05d",cz))
                ,begin = dmy_hms(begin)
                ,tz = toupper(str_trunc(tz, 3, side = "r", ellipsis = ""))
                ) %>%
        arrange(begin) 

        
names(wetbl) <- tolower(names(wetbl))

# Multiply out propety and crop damage variables to get full amounts
tmp <- as.numeric(str_extract(wetbl$damage_property,"^[^HKMB]*"))
mlt <- str_extract(wetbl$damage_property,"[HKMB]$")      
tmp2 <- if_else( mlt == "H",  10^2
         , if_else(mlt == "K",  10^3
                , if_else(mlt == "M", 10^6
                        , if_else(mlt == "B", 10^9
                                    , 10^0)
                  )
        )
, 10^0)

wetbl$damage_property <- tmp * tmp2
rm(list = c("mlt", "tmp", "tmp2"))

tmp <- as.numeric(str_extract(wetbl$damage_crops,"^[^HKMB]*"))
mlt <- str_extract(wetbl$damage_crops,"[HKMB]$")      
tmp2 <- as.numeric(if_else( mlt == "H",  10^2
                 , if_else(mlt == "K",  10^3
                           , if_else(mlt == "M", 10^6
                                     , if_else(mlt == "B", 10^9
                                               , 10^0)
                           )
                 )
), 10^0)
wetbl$damage_crops <- tmp * tmp2
rm(list = c("mlt", "tmp", "tmp2"))

head(wetbl)

event.damage <- transmute(wetbl
                        , id = id
                        , type
                        , begin
                        , year = as.factor(year(as.character.Date(begin)))
                        , injuries_direct 
                        , injuries_indirect 
                        , deaths_direct 
                        , deaths_indirect
                        , damage_crops
                        , damage_property
                        , fips = paste0(st,cz)
                        , fscale
                       ) %>%
                arrange(year, type) %>%
 mutate(wea_possible = as.logical(
               if_else(begin >= mdy("5/1/2012")
                       , true = "TRUE"
                       , false = "FALSE")
               )
)

```
# Selecting Event Types
With 52 event types, we are only going to focus on the most damaging ones.  I will look at the total of directly-caused, deaths and injuries.

```{r most_common, echo=-1, message=FALSE, warning=FALSE, cache=TRUE}
tmp <- select(event.damage
         , type
         , year
         , injuries_direct
         , injuries_indirect
         , deaths_direct
         , deaths_indirect
         , damage_property
         , damage_crops) %>%
        group_by(type) %>%
        summarize_at(c(3:8), c("sum", "mean"), na.rm = TRUE)

top_death <- top_n(tmp, 10, wt = deaths_direct_mean) %>%
        transmute(type = factor(type)
                  , direct = deaths_direct_sum
                  , indirect = deaths_indirect_sum) %>%
        arrange(desc(direct))

top_injury <- top_n(tmp, 10, wt = injuries_direct_mean) %>%
                transmute(type = factor(type)
                  , direct = injuries_direct_sum
                  , indirect = injuries_indirect_sum) %>%
        arrange(desc(direct))

top_damage <- top_n(tmp, 10, wt = damage_property_mean) %>%
               transmute(type = factor(type)
                  , property = damage_property_sum
                  , crops = damage_crops_sum) %>%
        arrange(desc(property))

print(top_death)
print(top_injury)
print(top_damage)

health.10 <- filter(event.damage
                    , type %in% top_death$type) %>%
             mutate(type = factor(type), ordered = TRUE)

health.mean.10 <- select(event.damage,1:8) %>%
        filter(type %in% top_death$type) %>%
        mutate(type = factor(type)) %>%
        group_by(year, type) %>%
        summarize_at(c(5:8), mean)

health.sum.10 <- select(event.damage,1:8) %>%
        filter(type %in% top_death$type) %>%
        mutate(type = factor(type)) %>%
        group_by(year, type) %>%
        summarize_at(c(5:8), sum)

hmean.long <- melt(health.mean.10, id.var = c("year","type")
                ,variable.name = "casualties")

hsum.long <- melt(health.sum.10, id.var = c("year","type")
                ,variable.name = "casualties")

```
# Data Plots
Wireless Emergency Alerts are designed to warn about imminent threats to life or property. By using text analyis tools on a database of all WEA messages sent since May of 2014, we can create a plot of frequency for the different types of messages. The National Weather Service (NWS) uses the WEA messages almost exclusively to warn about tornadoes and flash floods. 

<!-- Attach the text analysis of WEA messages -->
```{r text analysis of WEA messages, echo=TRUE, message=FALSE, warning=FALSE}

source('C:/Users/amsilverman/Box Sync/@Project Support/WEA_Analysis/NOAA Analysis/textMining.R')

```

The WEA messages are used for these types of weather events because

1. tornadoes and flash floods cause a very high proportion of weather-related casualties and 

1. tornadoes and flash floods are fast-moving and extremely localized events, both of which make WEA messages an ideal medium for warnings. 

# Effects of WEA on Weather-Related Injuries and Deaths

By plotting the casualties caused by similarly-sized tornadoes before and after the advent of WEA, we might see some effect of the warnings on population health. 

We will first use property damage as a proxy for event strength, as the 90-character WEA messages are more likely to help a person move to safety - the short notice that these alerts give are unlikely to give a recipient sufficient time to protect property before a tornado strikes.

```{r Plots, echo=TRUE, fig.width=7, message=FALSE, warning=FALSE}

# Total Casualties as a result of Weather Events#
 g <- ggplot(data = filter(hsum.long, casualties == "deaths_direct" | casualties == "injuries_direct"), 
            mapping = aes(type, value
                          , facets = casualties
                          , ylim(0,2))
            )

# 
g + geom_bar(stat = "identity", position = "stack") +
        facet_wrap(~casualties, ncol = 1, scales = "free_y") +
        scale_fill_brewer(palette = "Set3") + 
        theme_minimal() + 
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
        ggtitle("Casualties Due to Weather Events", subtitle = "By type 1997 - 2016, total")

# Yearly Casualties as a Direct consequence of Weather Events

g <- ggplot(data = filter(hsum.long, casualties == "deaths_direct" | casualties == "injuries_direct"),
            mapping = aes(year, value
                          , fill = type
                          , facets = casualties
                         # , ylim(0,2)
                          )
        )

# 
g + geom_bar(stat = "identity", position = "stack") +
        facet_wrap(~casualties, nrow = 4, scales = "free_y") + 
        scale_fill_brewer(palette = "Set3") +
        theme_minimal() + 
       theme(axis.text.x = element_text(angle = 90), complete = FALSE) + 
        ggtitle("Yearly Casualties due to Weather Events by Type", subtitle = "10 Most Common Types, 1997 - 2016") 

tornado <- filter(event.damage, type == "Tornado") %>%
                group_by(wea_possible)

m <- ggplot(data =  tornado %>% group_by(wea_possible)
            , mapping = aes(deaths_direct, damage_property
                            , facets = wea_possible)
        )
m + geom_point(na.rm = TRUE) +
        facet_grid(~wea_possible) +
        xlim(0,15) + ylim(0,10^8) + 
        geom_smooth(method = "lm", na.rm = TRUE, fullrange = TRUE) +
        scale_fill_brewer(palette = "Set3") +
        theme_minimal() +
        ggtitle("Tornado Deaths", subtitle = "Before and After WEA")

m <- ggplot(data =  tornado %>% group_by(wea_possible)
            , mapping = aes(injuries_direct, damage_property
                            , facets = wea_possible)
)
m + geom_point(na.rm = TRUE) +
        facet_wrap(~wea_possible) +
        xlim(0,15) + ylim(0,2*10^7) + 
        geom_smooth(method = "lm", na.rm = TRUE, fullrange = TRUE) +
        scale_fill_brewer(palette = "Set3") +
        theme_minimal() +
        ggtitle("Tornado Injuries", subtitle = "Before and After WEA")




```

## Strength of Tornado (Fscale), vs deaths and injuries 
Total deaths and injuries as functions of property damage seem marginally lower post-WEA, however we can also use the Enhanced Fujita Scale to directly compare the effects of similarly-sized tornadoes. 

Because there has only been one F5 tornado since WEA implementation in 2012 -- the Moore, Oklahoma tornado of May, 2013 which killed 24 and injured 207. Because of the small sample size, F5 tornadoes are omitted from this graph.
```{r tornado_casualties, echo=TRUE}
library(stringr)
library(reshape2)
tornado <- filter(event.damage, type == "Tornado") %>%
#        select(begin, year:wea_possible) %>%
        mutate(fscale = as.factor(
                str_extract(fscale, "[0-9]"))) %>%
        filter(!is.na(fscale), as.numeric(fscale) <= 5) %>%
        group_by(wea_possible)

big_tornado <- filter(tornado, as.numeric(fscale) > 4)

tornado_summary <- summarise_at(group_by(tornado
                                         , wea_possible
                                         , fscale)
                                , c("deaths_direct"
                                    , "injuries_direct"
                                    # , "deaths_indirect"
                                    # , "injuries_indirect"
                                    )
                                , mean
                                ) %>%        
        filter(as.numeric(fscale) < 6)


tornado_summary <- melt(tornado_summary
     , id.vars = c("wea_possible","fscale")
     , variable.name = "type"
     , value.name = "casualty")

n <- ggplot(data = tornado_summary
             , mapping = aes(fscale
                             , casualty
                             , facets = type
                             , fill = wea_possible
                             ))
n + geom_bar(stat = "identity", position = "dodge") + 
        facet_wrap(~type,nrow = 2 ,labeller = label_both, scales = "free") +
        labs(title = "Decreased Tornado Casualties Post-WEA"
             , x = "Tornado Strength (F Scale)"
             , y = "Mean Casualties") +
        scale_fill_brewer(palette = "Set3") + 
        theme_minimal()
        


```

